\normalfont \normalsize \textsc{Northwestern University} \\ [25pt] \horrule{0.5pt} \\[0.4cm] \huge Masters Thesis \\ \horrule{2pt} \\[0.5cm]

### 2016-08-28 Gradschool/thesis/final/bak/chap3.tex
expanded implementation chapter, merged bayesian statistics and GR chapter, relegated GR derivation to appendix

\documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size
\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\usepackage[adobe-utopia]{mathdesign}
%\usepackage{fourier} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
\usepackage[english]{babel} % English language/hyphenation
\usepackage{amsmath,amsfonts,amsthm} % Math packages
\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template
\usepackage{sectsty} % Allows customizing section commands

\usepackage{braket}
\usepackage{cancel}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage[mathscr]{euscript}

\allsectionsfont{\centering \normalfont\scshape} % Make all sections centered, the default font and small caps

\usepackage{fancyhdr} % Custom headers and footers
\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
\fancyfoot[L]{} % Empty left footer
\fancyfoot[C]{} % Empty center footer
\fancyfoot[R]{\thepage} % Page numbering for right footer
\renewcommand{\headrulewidth}{0pt} % Remove header underlines
\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
\setlength{\headheight}{13.6pt} % Customize the height of the header

\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)

\setlength\parindent{0pt} % Removes all indentation from paragraphs - comment this line for an assignment with lots of text

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height

\title{	
\normalfont \normalsize 
\textsc{Northwestern University} \\ [25pt] % Your us get downniversity, school and/or department name(s)
\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
\huge GPU Accelerated Parameter Estimation of Gravitational Waves from Binary Black Hole Mergers \\ % The assignment title
\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
}

\author{Brandon B. Miller, Masters Thesis} % Your name

\date{\normalsize\today} % Today's date or a custom date

\begin{document}

\maketitle % Print the title
\newpage
\section{Background and Motivation}

Gravitational waves are predicted as a mathematical consequence of the linearized Einstein Field Equations. The Einstein Field Equations themselves are a set of ten unique, coupled, nonlinear partial differential equations.

\subsection{Pieces of the Einstein Equation}
\subsubsection{Reimann Tensor}
The Reimann Tensor is a mathematical object that encapsulates the geometric \textit{curvature} of spacetime. The concept of a curved spacetime is fundamentally no different from the concept of a curved space, and from examining the behavior of vectors in such a space we can address both the purpose and properties of the Reimann tensor without departing from the intuitive picture of spheres and arrows.

\begin{figure}[h!]
\includegraphics[trim={2cm, 2cm, 2cm, 2cm}, clip]{spheredone.eps}
\caption{A visual depiction of the Reimann Curvature tensor. Parallel transporting a vector in the $\sigma$ direction both ways around the loop shown yields two different vectors. The orange vector $\vec{d}$ points between the tips. The magnitude of the components of this vector are $R_{\rho \sigma \mu \nu}$.}
\label{fig:sphere}
\end{figure}

The Reimann tensor is traditionally derived by examining the local behavior of geodesics subject to a metric tensor that describes curved space. Here we motivate the concept with a visual explanation of parallel transport. Consider a vector pointing in some arbitrary direction $\sigma$ somewhere on the sphere, as depicted in figure \ref{fig:sphere}. If we maintain the orientation of the vector relative to the surface of the sphere as we traverse the path, we note that the vector behaves differently depending on the order in which we choose to traverse the directions $\mu$ and $\nu$. This difference is representable by a vector that points between the two tips - one that is dependent on the initial direction $\sigma$ and both the transport directions $\mu$ and $\nu$. Let this vector be denoted $\vec{d}$. Then the Reimann tensor, $R_{\rho \sigma \mu \nu}$, is simply $d_{\rho}$, the $\rho$'th component of this difference vector. From this geometric picture some of the classical symmetries that are often introduced abreast the definition of the Reimann tensor are immediately intuitive, such as the skew symmetry depicted in figure \textbf{FIGURE HERE}.
Since it is somewhat unfair (worse - \textit{incorrect}) to characterize the entire geometry of the space with looping paths over large areas, the formal definition of the Reimann tensor is in fact the infinitesimal version of the concept above. Indeed, the concept of a single direction one may traverse is only valid in a small region of curved space, so the real Reimann tensor must describe the difference between two vectors as parallel transported around an infinitesimally small version of the original loop. It follows then how there must be a (possibly different) Reimann tensor at each point on the manifold in question. More remarkable however is the simplicity of the formal definition within this picture:

\begin{align}
R^{\rho}_{\sigma \mu \nu}\partial_{\rho} = \left(\nabla_{\mu}\nabla{\nu} - \nabla_{\nu}\nabla_{\mu}\right)\partial_{\sigma}
\end{align}

It should be noted that the above formula is true only modulo a term containing $\nabla_{[\mu, \nu]}$ except in the special case of coordinate vector fields, where it is zero. Although it does little to redeem it as a first explanation of the concept, this also illuminates why the Reimann tensor is sometimes described as encoding the non-commutativity of covariant derivatives on a manifold. Somewhat more subtle is the explicit role of the Reimann tensor in describing geodesic deviation. Crucial to a straightforward linearization of the Reimann tensor itself is the expression in terms of the Christoffel Symbols \textbf{AS DEFINED IN APPENDIX}

\begin{align}
R^{\rho}_{\sigma \mu \nu} = \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} + \Gamma^{\rho}_{\mu \lambda}\Gamma^{\lambda}_{\nu \sigma} - \Gamma^{\rho}_{\nu \lambda}\Gamma^{\lambda}_{\mu \sigma}
\end{align}

The christoffel symbols contain some information about the curvature of space as a function of the coordinates, and thus in the framework of linearized gravity products of Christoffel symbols are second order:

\begin{align}
R^{\rho}_{\sigma \mu \nu} = \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} 
\end{align}

If $x^{\rho}(\tau)$ are the components of a geodesic parameterized by $\tau$, then $\dot{x^{\rho}}(\tau) \equiv T^{\rho}$ are the components of a vector tangent to the geodesic. Suppose there is a family such nearby geodesics indexed by some other parameter, $s$, and that we can define \textit{deviation vector} $S^{\rho}(\tau) = \partial_{s}x^{rho}(s,\tau)$ which characterizes the separation between nearby geodesics. It can be shown that the second derivative of this vector with respect to $\tau$ is related to the Reimann tensor through the differential equation

\begin{align}
\ddot{S^{\rho}} = R^{\rho}_{\sigma \mu \nu} T^{\sigma} T^{\mu} S^{\nu}
\end{align}

\section{Linearized Gravity}

In linearized gravity we assume the spacetime is described by a flat spacetime $g_{\mu \nu}$ plus a small perturbation

\begin{align}
g_{\mu \nu} = \eta_{\mu \nu} + h_{\mu \nu}
\end{align}

Since the other pieces of the Einstein field equations are directly derived from the metric, they must be individually linearized to form a description of spacetime around weak sources. The standard treatment of linearized gravity assumes that to first order we may raise and lower indices using the Minkowski metric. If we tried to use the full form of equation \textbf{EQUATION}, we would have

\begin{align}
g^{\mu}_{\nu} &= g^{\mu \alpha}g_{\alpha \nu} \\
&= (\eta^{\mu \alpha} + h^{\mu \alpha}) (\eta_{\alpha \nu} + h_{\alpha \nu}) \\
&= \eta^{\mu \alpha}\eta_{\alpha \nu} + \eta^{\mu \alpha}h_{\alpha \nu} + h^{\mu \alpha}\eta_{\alpha \nu} + h^{\mu \alpha}h_{\alpha \nu}
\end{align}

The second two terms being second order in $h$, ther peturbation, are neglected in linearized gravity. We assume from here on that indices can be freely raised and lowered to first order using only the Minkowski metric $\eta_{\mu \nu}$.

\begin{align}
h^\mu_\nu = \eta^{\alpha \mu} h_{\alpha \nu} \ \  \text{and} \ \ h^{\mu \nu} = \eta^{\alpha \mu}\eta^{\beta \nu}h_{\alpha \beta}
\end{align}

The next natural line of logic to pursue is to write down the Christoffel symbols for such a spacetime

\begin{align}
\Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2}g^{\nu \mu}\left[\partial_{\beta}g_{\alpha \nu} + \partial_{\alpha}g_{\beta \nu} - \partial_{\nu}g_{\alpha \beta}\right] \\
&= \frac{1}{2}\left(\eta^{\mu \nu} + h^{\mu \nu}\right)\left[\partial_{\beta}g_{\alpha \nu} + \partial_{\alpha}g_{\beta \nu} - \partial_{\nu}g_{\alpha \beta}\right]
\end{align}

Inserting the expanded form of the metric into the term within the brackets gives us six terms times two at the front, for twelve total. Thankfully, three correspond to the standard flat-space Christoffel symbols which can be made zero through a choice of coordinates (in fact, orthonormal Cartesian coordinates), three involve derivatives of $\eta_{\mu \nu}$ which are zero, and three are second order in $h_{\mu \nu}$, which we assume are zero. This leaves us with three:

\begin{align}
\Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2}\eta^{\mu \nu}\left[\partial_{\beta}h_{\alpha \nu} + \partial_{\alpha} h_{\beta \nu} - \partial_{\nu}h_{\alpha \beta}\right] \\
&= \frac{1}{2} \left[\partial_{\beta} h^{\mu}_{\alpha} + \partial_{\alpha}h_{\beta}^{\mu} - \partial^{\mu}h_{\alpha \beta}\right] 
\end{align}

Where in the last term $\partial^{\mu} = \eta^{\mu \nu}\partial_{\nu}$ was used to raise the index on the actual partial derivative operator itself. It should be noted at this point that since we neglect terms that are second order in the Christoffel symbols themselves, in linearized gravity, the covariant derivative of a metric perturbation is equivalent to the normal partial derivate. The same assumption allows for simplification of the Reimann and Ricci tensors:

\begin{align}
R^{\rho}_{\mu \sigma \nu} &= \partial_{\sigma}\Gamma^{\rho}_{\mu \nu} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} \\
R_{\mu \nu} = R^{\rho}_{\mu \rho \nu} &= \partial_{\rho}\Gamma^{\rho}_{\nu \mu} - \partial_{\mu}\Gamma^{\rho}_{\mu \rho}
\end{align}

Plugging in equation \textbf{EQUATION} we have

\begin{align}
R_{\mu \nu} & = \frac{1}{2}\partial_{\alpha}\left[\partial_{\nu}h^{\alpha}_{\mu} + \partial_{\mu}h^{\alpha}_{\nu} - \partial^{\alpha}h_{\mu \nu}\right] - \frac{1}{2}\partial_{\nu}\left[\partial_{\alpha}h^{\alpha}_{\mu} + \partial_{\mu}h^{\alpha}_{\alpha} - \partial^{\alpha}h_{\mu \alpha}\right] \\
&= \frac{1}{2}\left[ \cancel{\partial_{\alpha}\partial_{\nu}h^{\alpha}_{\mu}} + \partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \cancel{\partial_{\nu}\partial_{\alpha}h^{\alpha}_{\mu}} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} \right] \\
&= \frac{1}{2}\left[\partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha}\right]
\end{align}

The quantity $h^{\alpha}_{\alpha}$ is the trace of the pertubation metric and represents, in a sense, an overall measure of the \textit{strength} or \textit{amplitude} of the perturbation, this is often called just $h$. The term $\partial^{\alpha}\partial_{\alpha}$ is an inner product that when formed through the Minkowski metric becomes the wave operator, $\Box = -\partial_{t}^{2} + \nabla^2$. The Ricci tensor is obtained by contracting the Reimann tensor over the two free indices:

\begin{align}
R \equiv \eta^{\mu \nu}R_{\mu \nu} &= \frac{1}{2}\left[\partial_{\alpha}\partial^{\mu}h_{\mu}^{\alpha} + \partial_{\mu} \partial^{\alpha}h^{\mu}_{\alpha} - \Box h - \Box h\right]
\end{align}

Since $\alpha$ and $\mu$ are summed indices, the first two terms are the same. Dropping the $\frac{1}{2}$, the Ricci scalar curvature is

\begin{align}
R = \partial_{\alpha}\partial_{\mu}h^{\mu \alpha} - \Box h
\end{align}

Many authors choose to simplify the expression for the Ricci Tensor by defining the quantity

\begin{align}
V_{\nu} = \partial_{\alpha}h^{\alpha}_{\nu} - \frac{1}{2}\partial_{\nu}h
\end{align}

With which the first two terms of the linearized Ricci tensor can be reorganized as

\begin{align}
\partial_{\mu}\partial_{\alpha}h_{\nu}^{\alpha} - \partial_{\mu}\partial_{\nu}h &= \partial_{\mu}\left[\partial_{\alpha}h_{\nu}^{\alpha} - \partial_{\nu}h\right] \\
&= \partial_{\mu}V_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h
\end{align}

Which upon substitution into the \textbf{EQUATION} for $R_{\mu \nu}$ yields

\begin{align}
R_{\mu \nu} &= \frac{1}{2}\left[\partial_{\mu}V_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h + \partial^{\alpha}\partial_{\nu}h_{\mu \alpha} - \Box h_{\mu \nu}\right]
\end{align}

The two unsimplified terms form the quantity $\partial_{\nu}\left[\partial^{\alpha}h_{\mu \alpha} - \frac{1}{2}\partial_{\mu}h\right]$ which we recognize as $\partial_{\nu}V_{\mu}$, leaving us with

\begin{align}
R_{\mu \nu} = \frac{1}{2}\left[\partial_{\mu}V_{\nu} + \partial_{\nu}V_{\mu} - \Box h_{\mu \nu}\right] 
\end{align}

Setting the right hand side of this equation gives us the linearized, vacuum Einstein Equation $R_{\mu \nu} = 0$:

\begin{align}
\frac{1}{2}\left[\partial_{\mu}V_{\nu} + \partial_{\nu}V_{\mu} - \Box h_{\mu \nu}\right] = 0 
\end{align}

$h_{\mu \nu}$ is a two index object in a four dimensional spacetime, which in the general case would present with sixteen independent components each of which would need to be explicitly determined to form a complete picture of the physics. However in our specific case it will turn out that there are far fewer real independent components then this, which can be observed by chiseling away at the above expression in the following manner. Firstly, the metric of general relativity is taken to be symmetric, because if we assume that the tensor $dx^{\mu}dx^{\nu}$ is symmetric (i.e. $dx^{\nu}$ and $dx^{\nu}$ commute) then only the symmetric part of any $g_{\mu \nu}$ contributes when the quantity $g_{\mu \nu}dx^{\mu}dx^{\nu}$ is computed. Thus we have

\begin{align}
h_{\mu \nu} = h_{\nu \mu}
\end{align}

Which, for a sixteen component tensor, makes six of the components redundant. This leaves us with ten. Furthermore, there is \textit{gauge freedom} present in the linearized vacuum Einstein equation. This is best illustrated with reference to the equations of classical electrodynamics, which describe the motion of charged particles subject to the electric and magnetic vector potentials, $V$ and $\vec{A}$ respectively. It is somewhat easily demonstrable that the same equations of motion for electrically charged particles are unchanged when $V$ and $\vec{A}$ are subject to transformations of the form

\begin{align}
\vec{A} &\rightarrow \vec{A} + \nabla \Psi \\
V &\rightarrow V -\frac{\partial \Psi}{\partial t}
\end{align}

The function $Psi$ is known as the \textit{gauge function} and it can be absolutely anything we want, as long as we adjust the potentials accordingly. Within such a framework, one can impose certain conditions on the potentials that simplify the mathematics, such as requiring that

\begin{align}
\nabla \cdot \vec{A} + \frac{1}{c}\frac{\partial V}{\partial t} &= 0 \\
\rightarrow \partial^{\mu}A_{\mu} &= 0
\end{align}

It is somewhat counterintuitive why imposing this particular condition on $\vec{A}$ and $V$ themselves is equivalent to transforming the potentials via the gauge function, but indeed it is, it is just hidden! While we have made no reference to $\Psi$, we have \textit{implicitly} assumed that there is such a $\Psi$ that will \textit{make} the restriction on $\vec{A}$ and $V$ true. It actually is there inside the $V$ and $\vec{A}$ written in equation \textbf{EQUATION}. We do not have to state \textit{what it is}, as long as we know that it exists. A proof that there legitimately is such a gauge function exists but is out of the scope of this outline. 
There is a one-to-one correspondance of this formalism with the equivalent gauge in general relativity, which is known as the Lorentz gauge. In the same sense as equation \textbf{EQUATION}, it turns out that the equations of motion for massive particles in a slightly perturbed spacetime are invariant under coordinate transformations of the form

\begin{align}
x^{\mu} \rightarrow x'^{\mu} - \xi^{\mu}
\end{align}

In particular, we can show that there exists a coordinate transformation such that

\begin{align}
\partial_{\mu}V_{\nu} = \partial_{\mu}\partial_{\alpha}h^{\alpha}_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h = 0
\end{align}

\textbf{PUT PROOF HERE}

\begin{align}
\text{PUT PROOF HERE}
\end{align}

Thus coordinate transformations that obey $\Box \xi = 0$ will allow us to impose the Lorentz gauge. Within this gauge, the linearized, vacuum Einstein equation becomes simply

\begin{align}\label{eq:flteinst}
\Box h_{\mu \nu} = 0
\end{align}

This is a four dimensional wave equation for the remaining components of $h_{\mu \nu}$. Moreover, we have shown that under coordinate transformations that do not change the underlying physics, there exists an explicit relationship between components of the metric given by equation \textbf{EQUATION}. In the same way that symmetry shows that six of the metric components to be functions of the others, The index $\nu$ indicates that in four dimensions there are four additional explicit relationships between metric components that correspond to a loss of four additional degrees of freedom. This leaves us with six.

In flat space, equation \ref{eq:flteinst} for some arbitrary function $f(x)$ in place of the $h_{\mu \nu}$ would read

\begin{align}
\Box f(x) = \eta_{\alpha \beta} \frac{\partial^2 f}{\partial x^{\alpha} \partial x^{\beta}} &= 0\\
-\frac{\partial^2 f}{\partial t^2} + \nabla^2f &= 0
\end{align}

This is the general form of the time dependent wave equation whose solutions take the form

\begin{align}
f(x) = a e^{i \mathbf{k} \cdot \mathbf{x}}
\end{align}

Where $\mathbf{x}$ is the standard positon four-vector and $\mathbf{k}$ is the four-wave-vector. Expanding the inner product in the exponential again using the Minkowski metric as $-k^t t + \vec{k}\cdot \vec{x}$ and plugging the result into equation \ref{eq:flteinst} we have

\begin{align}
\text{PUT EQ HERE}
\end{align}

\section{Bayesian Statistics}
We select a Bayesian posterior probability as our main analysis technique as its ability to quantify and propagate uncertainties in parameter estimates is well documented by \textbf{CITATION} and \textbf{PROBABLY OTHER CITATIONS}. Bayes' Theorem follows from the symmetric definition of conditional probability depicted visually in figure \ref{fig:distrib}. Let the joint probability evaluated at any point $(x,y)$ in the plane be denoted $P(x,y)$. We may take a cross section through the joint probability distribution along any direction we choose, let us call such a cross section parallel to the $x$ axis the conditional probability of $x$ given $y$, and denote it $P(x|y)$. A perpandicular cross section would simply be $P(y|x)$. We see that we can compute $P(x,y)$ in two equivalent ways: scaling $P(x|y)$ by $P(y)$ or by scaling $P(y|x)$ by $P(x)$. Thus we have that

\begin{figure}[h!]
\center
\includegraphics[trim={1cm, 1cm, 1cm, 1cm}, clip]{biviariate_norm_done.eps}
\caption{A joint probability distribution $P(x,y)$. The contours parallel with the $x$ direction represent different $P(y|x)$.}
\label{fig:sphere}
\end{figure}

\begin{align} \label{eq:bayes}
P(x|y)P(y) &= P(y|x)P(x) \\
P(x|y) &= \frac{P(y|x)P(x)}{P(y)}
\end{align}

Which is Bayes' theorem. The denominator of the expression on the right hand side is often rewritten as a sum over all possible situations in which $y$ is an outcome. Reading $\neg x$ as "not x", we have

\begin{align}\label{eq:continuous}
P(y) &= P(y|x)P(x) + P(y|\neg x)P(\neg x) \\
&= \sum_{k=1}^K P(y|x_k)P(x_k) 
\end{align}

\clearpage 
This allows us to rewrite the posterior probability as

\begin{align}
P(x|y) = \frac{P(y|x) P(x)}{\sum_{k=1}^K P(y|x_k)P(x_k)}
\end{align}

Multiplying the right hand side of this equation by the quantity $\frac{1}{P(x)P(y|\neg x)}$ one has

\begin{align} \label{eq:lratio}
P(x|y) &= \frac{\Lambda(x|y)}{\Lambda(x|y) + \frac{P(\neg x)}{P(x)}}
\end{align}

Where we have used the \textit{likelihood ratio} $\Lambda(x|y) \equiv \frac{P(y|x)}{P(y|\neg x)}$ to simplify the expression. The likelihood ratio has a simple interpretation, it is simply the ratio of the probability of event $y$ occuring in the case event $x$ occurs and the probability of event $y$ occuring given that event $x$ \textit{does not} occur. In fact, this form of Bayes theorem indicates that the posterior probability for $x$ given $y$ is directly proportional to the likelihood ratio. This is particularly important in the Bayesian treatment of gravitational waves where the central question for study is the probability of observing the data in question given that there may or may not be a signal present.

\subsection{Signal Processing}
Noise in the LIGO detectors is modeled as \textit{Gaussian Noise}, which describes a length $N$ set of time series samples $\{x_j\}$ with the following probability of observation:

\begin{align}
P(\{x_j\}) = \left[\frac{1}{\sqrt{2 \pi} \sigma}\right]^N e^{-\frac{1}{2 \sigma^2} \sum_{j = 0}^{N - 1}x_j^2}
\end{align}

If we allow the $\{x_j\}$ to be weighted by the noise sensitivity in frequency space (also known as the \textit{power spectral density}) $S(f) = \sum_{k=0}^K n^*(f_k)n(f_k) \equiv \braket{n(f)|n(f)}$ then we may write the probability density for any Gaussian noise process as

\begin{align}
P_x(\{x(t)\}) \propto e^{-\frac{1}{2}\braket{x(t)|x(t)}}
\end{align}

For gravitational wave signal processing we are interested in evaluating the equivalent of the likelihood ratio  in equation \ref{eq:lratio} for the null and signal present hypotheses, $\mathcal{H}_0$ and $\mathcal{H}_1$ respectively. This is

\begin{align}
\Lambda(\mathcal{H}_1|d) &= \frac{P(d|\mathcal{H}_1)}{P(d|\mathcal{H}_0)}
\end{align}

In the above equation $\mathcal{H}_0$ is in direct correspondence with $\neg x$ and $\mathcal{H}_1$ is in direct correspondence with $x$. If there is no signal present in the data, then the data and noise are equal, meaning

\begin{align}
d(t) &= n(t) \\
P(d(t)|\mathcal{H}_0) &\propto e^{-\frac{1}{2}\braket{d(t)|d(t)}}
\end{align}

Whereas if there is a signal present in the data, then the noise is the \textit{residual} signal after the gravitational wave potion of the signal $h(t)$ is subtracted from the total signal $d(t)$:

\begin{align}
d(t) &= n(t) + h(t) \\
p(d(t)|\mathcal{H}_1) &\propto e^{-\frac{1}{2}\braket{d(t) - h(t)|d(t) - h(t)}}
\end{align}

So the likelihood ratio for gravitational wave signal processing is exactly equal to

\begin{align}
\Lambda(\mathcal{H}_1|d(t)) &= \frac{e^{-\frac{1}{2}\braket{h(t)-d(t)|h(t)-d(t)}}}{e^{-\frac{1}{2}\braket{d(t)|d(t)}}}
\end{align}

By equation \ref{eq:bayes}  the posterior probability of a gravitational wave signal with parameters $\mu$ is given by

\begin{align}\label{eq:parampos}
P[\vec{\mu}, \mathcal{H}_1|d(t)] = \frac{P[d(t)|\vec{\mu}, \mathcal{H}_1]P(\vec{\mu})}{P[d(t), \mathcal{H}_1]}
\end{align}

Extending equation \ref{eq:continuous} to the case of a continuous state space we have for the denominator

\begin{align}
P[d(t), \mathcal{H}_1] &= \int_{\Omega}P[d(t)|\vec{\mu}, \mathcal{H}_1]P(\vec{\mu})d\vec{\mu}
\end{align}

Where $\Omega$ indicates integration over the entire probabilistic state space. Dividing both the top and bottom of the right hand side of equation \ref{eq:parampos} by $P(d(t)|\vec{\mu}, \mathcal{H}_1)$ gives us

\begin{align}\label{eq:bayeslikely}
P(\vec{\mu}, \mathcal{H}_1|d(t)) = \frac{\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)P(\vec{\mu})}{\int_{\Omega}\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)P(\vec{\mu})d\vec{\mu}}
\end{align}

Note that the integral in the denominator may be factored into components that are functions of independent groups of parameters. For the purposes of gravitational wave data analysis it is factored into intrinsic and extrinsic parameters, the former (masses, spins) being responsible for the actual orbital dynamics of the system, and the latter (distance, inclination, right ascension, declination) depending only on the nature of the observer relative to the source. We label these sets of paramters $\vec{\lambda}$ and $\vec{\theta}$ respectively. The denominator of equation \ref{eq:bayeslikely} is thus factorizable as

\begin{align}
\int_{\Omega}\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)d\vec{\mu} &= \int_{\Omega}\Lambda(d(t)|\vec{\lambda}, \mathcal{H}_1)P(\vec{\lambda})d\vec{\lambda}\int_{\Omega}\Lambda(d(t)|\vec{\theta}, \mathcal{H}_1)P(\vec{\theta})d\vec{\theta} 
\end{align}

Note that while evaluating the likelihood for a single set of parameters in the numerator of the above equation may not be any serious challenge, computing the integral in the denominator is. Many schemes \textbf{CITE BEN} have been developed to tackle this particular problem and generally involve simulating the posterior through MCMC sampling. These techniques are highly developed and at the time of writing can provide converged results in \textbf{SOME TIMESCALE}, however they are limited in their parallelism due fundamental serialization of the algorithm. Cutting edge implementations involving parallel tempering and some forms of ensemble MCMC have shown promising results, however none can take advantage of the many thousands of cores availble across the LIGO data grid. A new and more embarassingly parallel scheme is needed.

\subsection{Harmonic Mode Decomposition}

For a single detector $k$ We can write our gravitational wave signal $h_k(t) = h_{+,k}(t) - ih_{\times, k}(t)$ as a sum over products of harmonic mode strain values $h_{l,m}(\vec{\lambda})$ and corresponding spin-weighted spherical harmonics, scaled by the reference distance and antenna factor:

\begin{align}
h_k(t) = \frac{D_r}{D}F_k(\theta)\sum_{(l,m)}h_{k, (l,m)}(t, \vec{\lambda})Y_{(l,m)}(\vec{\theta})
\end{align}

Where $D_r$ is the luminosity reference distance to the binary. We substitute this expression for the signal into the expression for the likelihood ratio we find, taking the logarithm, that

\begin{align}
\mathcal{L} &= \prod_k e^{-\frac{1}{2}\braket{d - h_k|d-h_k} + \braket{d|d}} \\
\ln\mathcal{L} &= -\frac{1}{2}\sum_{k}\left[\braket{d - h_k|d-h_k} + \braket{d|d}\right] \\
&= \frac{1}{2}\sum_{k}\left[-\cancel{\braket{d|d}}+\braket{d|h_k}+\braket{h_k|d}+\braket{h_k|h_k} + \cancel{\braket{d|d}}\right] \\
\end{align}

Now, since for complex vector spaces $\braket{a|b} = \braket{b|a}*$, we have that $\braket{d|h_k} + \braket{h_k|d} = 2\Re\braket{h_k|d}$. Then,

\begin{align}
\ln\mathcal{L} &= \frac{1}{2}\sum_{k}\left[2\braket{h_k|d} + \braket{h_k|h_k}\right] \\
&= \frac{1}{2} \sum_{k}\Re\left[\braket{2\frac{D_r}{D}F_k(\vec{\theta})\sum_{(l,m)}h_{k,(l,m)}(\lambda)Y_{(l,m)}(\vec{\theta})|d}\right] \\
&+ \left[\frac{D_r}{D}\right]^2F_k^2(\vec{\theta})\sum_{(l,m),(l',m')}h^*_{(l,m)}(\vec{\lambda})h_{(l',m')}(\vec{\lambda})Y^*_{(l,m)}(\vec{\theta})Y_{(l',m')}(\vec{\theta})
\end{align}

Now, for any complex number $z = a + ib$, we have that $z*z + \Re(z^2) = 2z*z$, so $z*z = \frac{1}{2}(z*z + \Re z^2)$. Then the double sum over $l,m$ pairs becomes

The defining feature of this form of the likelihood is the separation between intrinsic and extrinsic parameters. The intrinsic parameters of the binary exist only within the $h_{k,(l,m)}$, whereas the parameters that depend on the observer are contained entirely within the antenna factor and spherical harmonics. Work by \textbf{CITE CHRIS AND RICHARD} has shown that this can be exploited to accelerate computation of the likelihood by precomputing the orbital dynamics of the binary and marginalizing out all of the extrinsic parameters with a Monte Carlo integration.

@@

\section{Results and Prospects for Low Latency Electromagnetic Followup}
\begin{figure}[H!]
\hspace{-5.0cm}
\includegraphics[scale=0.57]{speedup.eps}
\end{figure}

The dawn of gravitational wave astronomy in the 2015-2016 era has touched off a revolution in our ability to observe and classify astronomical phenomena. As with any entirely novel method of gleaning information from distant sources, development of methods to process and decipher such data has been explosive. Pure gravitational wave astrophysics has exciting prospects for the future, but the possibility of advancing other classical forms of astrophysics through their interplay with these newer methods is equally compelling. Significant research \textbf{CITATION} in th epast several years has been designed specifically to connect electromagnetic astronomy to gravitational wave observations, in particular the study of low-latency electromagnetic follow-up to gravitatioanl wave triggers. A central point of focus in these studies is the opportunity to give electromagnetic observers early warning on possible short duration transient events such as gamma ray bursts or longer duration klonovae. Such early warning could provide valuable insights as to the central engine of such gamma ray bursts and confirm recent models of these processes. However, electromagnetic observational resources are limited and expensive to operate and such it is imperative that any communication to such resources in the context of EM followup of GW triggers must be rapid and accurate to avoid wasting time and money. To that end it is desireable to develop existing parameter estimation codes to such a point that they can produces fast and correct results to reliably indicate potential sources of interest while minimizing false positives. The main goal of this work is to push this envelope and reduce the end-to-end runtime of parallel PE codes (specifically RapidPE) for the express purpose of low latency electromagnetic follow-up.

\subsection{Gamma Ray Bursts}
Gamma ray bursts (GRBs) are short duration transient events thought to occur during or shortly after the merger of compact binary systems, specifically those which include a single black hole and one neutron star. The presence of strong radiative emission from the poles of black holes has been confirmed by \textbf{CITATION}, and while not yet directly observed, the interaction between such polar jets of radiation and the resulting post-merger NS remnant is a primary candidate for the generation of GRBs. The process is depicted in figure \textbf{FIGURE} in which an infalling NS is torn apart by the BH counterpart, leaving a remnant disk of energetic matter in orbit around the black hole. The radiation eminated from the poles of the BH impacts this disk and generates gamma rays, this is known as an internal shock. As the resulting cloud of matter ploughs into the surrounding interstellar medium, it slows, leaving a lower density region in its wake. Subsequent waves of matter may propagate through this region at a higher velocity then the initial blast wave, catching up and ramming into the back of the initial wave, producing further shocks.
It is not yet explicitly confirmed that this is the exact process by which GRBs are generated. It is generally accepted that to confirm this hypothesis we would need to observe direct coincidence between a GW trigger localized to some region of the sky and a GRB from the same direction. These phenomena have been demonstrated in numerical simulations by \textbf{CITATION} and are highly dependent on initial system parameters. In particular the currently unknown commplete NS equation of state is known to play a pivotal role in braketing the mass and mass ratio ranges associated with GRB progenitors \textbf{CITATION}. We now attempt to demonstrate the efficacy of the GPU based version of RapidPE for estimating these parammeters quickly and accurately. We use rate predictions from \textbf{CITATION} to restrict the mass range to be studied to fall between \textbf{SOME RANGE}. These bounds corespond to a chirp mass range of \textbf{WHAT RANGE}. From this we can determine the expected time that the signal will spend in the Advanced LIGO band. In figure \textbf{FIGURE}, we have visualized this range in \textbf{SOME PLANE}. Using the results of our code we find that anadditional \textbf{SOME PERCENTAGE} of expected NSBH sytems expected to be picked up by Advanced LIGO are now candidates for joint GW-EM astronomy.

\begin{figure}[H!]
\hspace{-0.5cm}
\includegraphics[scale=0.35]{gamray.png}
\end{figure}

>>>>

%% This is an example first chapter.  You should put chapter/appendix that you
%% write into a separate file, and add a line \include{yourfilename} to
%% main.tex, where `yourfilename.tex' is the name of the chapter/appendix file.
%% You can process specific files by typing their names in at the 
%% \files=
%% prompt when you run the file main.tex through LaTeX.
\chapter{Implementation}

@@
